Modeling Intervention Strategies for United States TB Control

### 2014-08-03 in_progress/presentations/paper/draft.tex
Added some scripts for the paper and worked on the paper

% arara: pdflatex
% arara: bibtex
% arara: pdflatex
% arara: pdflatex
\documentclass{amsart}
\usepackage[pdftex]{graphicx}
\usepackage{enumerate,placeins}
\usepackage{color}
\graphicspath{{figures/}}

\newcommand{\TODO}[1]{\hfill\\{\huge \color{red} TO DO:} #1 \hfill \\}

@@

{\huge \color{red} List of results}
Plan:
\begin{enumerate}
  \item Intervention Analysis: 
    \begin{enumerate}
      \item Reducing Entering LTBI
      \item Cutting Transmission
      \item Reducing Progression Rate
    \end{enumerate}
  \item Qualitative Strategy Conclusions:
    \begin{enumerate}
      \item Sourced Incidence \& Cost
      \item Plateauing Effect
      \item Extremal values for entering LTBI reduction
      \item Linearity in Cost/Incidence for entering LTBI reduction
      \item CPCA data
    \end{enumerate}
  \item Statistical Verification/Agent Based Feasibility
    \begin{enumerate}
      \item Normal Distributions
      \item Basic Reproductive Numbers
      \item Agent based running time \& Extendibility. 
    \end{enumerate}
\end{enumerate}
Ideas:
\begin{itemize}
  \item Plateauing Effect
  \item Intervention Analysis
  \item Statistical Analysis of Hill Model
  \item Agent-Based Feasibility
  \item Theoretical/Experimental justification of basic reproduction number
  \item Overall qualitative strategy conclusions. 
\end{itemize}

\section{Abstract}
An epidemiological and economic model for tuberculosis in the United States
 was developed, extending a model developed in 2012 by the Centers
for Disease Control and Prevention. Sensitivity analysis of this model reveals
that the percentage of foreign born immigrants with LTBI is a highly influential
parameter. Various intervention strategies, particularly those that reduced LTBI
among foreign-born arrivals, were evaluated, and cost per case averted estimates
were found.  Additionally, a population-level, stochastic, agent-based model was
built. This model demonstrated the feasibility of using agent-based modeling in
the context of minimal population heterogeneity and provided statistical
characterization of the results of the compartmental model. The distributions
of the final population sizes and incidence rates projected by the stochastic model
were found to be approximately normal, with means centered at the results of
compartmental model.

\section{Introduction}
Epidemiological models allow public health professionals to predict and analyze
disease dynamics and intervention effectiveness. The most common examples of
such models are detereministic, compartmental, differential equation models. In
these models, the population is split between several possible health states, or
compartments, and a system of differential equations governs population flow
between the various compartments. In 2012, Hill, Becerra, and Castro implemented
a compartmental differential equaiton model of tuberculosis (TB) in the United
States (US).  Their model characterized five health states across two
subpopulations, US-born (USB) and foreign-born (FB), for a total of 10
compartments. They used this model to evaluate several possible intervention
strategies and ultimately concluded that though increasing LTBI treatment was an
effective intervention strategy, the US was unlikely to meet their stated goal
of elimination of TB in the US by 2100. Here, we present an extended Hill model
with additional tracking capabilities, such that it can now report
further granularity in the disease dynamics. Furthermore, this model also tracks
economic data in order to project the US health care system (HCS) costs due to
TB given our current policy, as well as in the context of various interventions.

{\huge \color{red} Fix this paragraph. Its not the best. \textbackslash/}

Though compartmental, differential equation models are the most common strategy
when modeling large-scale diseases at the population level, an alternative
approach is to use an agent-based model. Agent
based models provide a more nuanced view of disease spread as they provide a
more biologically accurate modeling framework; however, they are often
considered to be computationally infeasible at population level. In order to
challenge this assumption of infeasibility and provide statistical verification
of the results of the compartmental model, we implemented a stochastic,
agent-based model of US TB dynamics at the population level.

\section{Background}
\subsection{Tuberculosis} x 
\TODO{Use TB instead of Tuberculosis maybe?}
Tuberculosis is an infectious bacterial disease that primarily affects the lungs,
with common symptoms including severe coughing, fevers, and chest pain.
Tuberculosis is a major global health concern; The World Health Organization
estimates that 8.6 million developed TB and 1.3 million people died from
the disease in 2012. Tuberculosis is an airborne disease and is typically
transmitted from person-to-person by coughing, sneezing, and the exchange of
other respiratory fluids.  Infected individuals first enter a latent phase,
during which the disease causing agent, \emph{Mycobacterium tuberculosis}, is
present in their bodies, but the individuals remain asymptomatic and
noncontagious. During this phase, individuals are said to have latent TB
infection (LTBI). Eventually, some number of individuals with LTBI progress into
what is known as the active phase of the disease, or active TB. Individuals with
an active TB infection are symptomatic and contagious. The lifetime risk of
progressing from Latent TB Infection (LTBI) to active TB is approximately
5-10\%, although individuals with LTBI have the greatest chance of developing
Active TB during the first two years immediately following infection, and if
they have a suppressed immune system, such as in the case of a comorbid HIV
infection. Cases of Active TB may be classified as pulmonary ($\sim$80\% of
cases) or non-pulmonary ($\sim$20\% of cases), depending upon the biological
location of infection. In the Hill model, pulmonary Active TB is treated as
contagious and non-pulmonary Active TB is treated as non-contagious, which is an
assumption we also use here.

In the United States, 9,951 new cases of TB were reported in 2012 by the Centers
for Disease Control. The disease burden varied for different subpopulations in
the U.S., and, in particular, the incidence rate of TB was 11.5 times higher for
the FB population than the USB population.  For the
past 20 years, the overall incidence of TB in the U.S. has declined each year,
but in 2012 the annual incidence level of 32 new cases/million remains above the
desired goal of the $<$1 new case/million, which is the target for disease
elimination set by the Centers for Disease Control.

\TODO{}citations: CDC Tuberuclosis 2013 Report - http://www.cdc.gov/mmwr/preview/mmwrhtml/mm6211a2.htm
WHO Global Tuberculosis Report 2013 - http://www.who.int/tb/publications/global\_report/en/

\subsection{The Hill Model}
A flowchart representation of the Hill Model is shown in
Figure~\ref{fig:hillModelSchematic}. Each compartment represents a different
possible TB health state for every US-born or FB individual, and
arrows between different compartments represent possible transitions between
states. Individuals also leave the model from all compartments due to natural
death, which is left out of the figure for clarity.

\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.25]{figures/HillModelFlowChart.png}
  \end{center}
  \caption{Schematic of the Hill Model.}
  \label{fig:hillModelSchematic}
\end{figure}

The majority of USB and FB individuals fall into the Susceptible (S) category,
which includes everyone who is uninfected and has not been exposed to TB, or
those who have been successfully cured of the disease. After exposure to an
individual with TB, a person in the Susceptible compartment can develop Latent
TB Infection (LTBI). Latently infected individuals are asymptomatic and
non-infectious, but have some risk of developing active TB infection over time.
To reflect the fact that real LTBI patients have a much higher risk of
developing active TB within two years of exposure, the Hill Model splits the
LTBI compartment into Acute LTBI (fast progressors) and Chronic LTBI (slow
progressors). Accordingly, individuals in the Acute LTBI compartment have a
higher risk of developing active TB than those in the Chronic LTBI compartment.
Individuals in the Chronic LTBI compartment may also be exogenously re-infected
and transition to the Acute LTBI compartment.

Latently infected individuals may progress to one of two active TB states: Infectious TB (I) or 
Non-Infectious TB (J).  Individuals in both compartments have an increased risk of death from
active TB infection, but only individuals in the Infectious TB compartment are contagious.  
In addition, individuals in all of the infected compartments (F, L, I, J) may be treated or self-cure
themselves of their respective TB health condition.  However, in the model, treatment or self-cure from TB
does not grant immunity, and all healthy individuals are grouped in the Susceptible compartment
and may be re-infected at a later time.\\

\section{Methods}
\subsection{Basic Structure}
Both the Hill Model (Basic Hill Model) and our extended Model (Extended Hill
Model) were implemented in \texttt{R} as a system of differential equations,
which were solved via the $\texttt{lsoda}$ routine. The systems of differential
equations and relevant constants used to capture basic disease dynamics in the
Basic Hill Model are shown in Figures~\ref{fig:hillEquations}
and~\ref{fig:basicHillConstants}, though a full analysis of the model is better
suited for \TODO{<+Cite Hill+>}. \TODO{Add equations for $\lambda$. I don't have
a pdf of Hill, apparently, though, so I can't.}  In the Extended Hill Model,
each flow rate equation from the Basic Hill Model was separated into the
components that tracked specific population migrations (i.e., terms that dictate
flow rates from one compartment to another) and these were tracked separately.
In other words, rather than tracking net flow into various compartments, the
Extended Hill Model tracks flow between all pairs of compartments separately.
Furthermore, these equations were each also supplemented to estimate their
corresponding US HCS costs. A relevant subset of the Extended Hill equations are
detailed in the appendix, section~\ref{sec:extendedHillEqs}. Historical data
from 2000-2008 was used to initialize the model, which was then run up to the
year 2100.

\begin{figure}
  \begin{align*}
    \dot{S}_0 &= \rho(N_0+N_1) + \sigma_0^FF_0 + \sigma^LL_0 
                +\varphi_0(I_0+J_0) - \lambda_0 S_0 - \mu_0 S_0\\
    \dot{F}_0 &= p\lambda_0S_0 + xp\lambda_0L_0 - (\mu_0+\nu^F+\sigma_0^F)F_0\\
    \dot{L}_0 &= (1-p)\lambda_0S_0 - xp\lambda_0L_0 
                -(\mu_0+\nu_0^L+\sigma^L)L_0\\
    \dot{I}_0 &= q(\nu^FF_0 + \nu_0^LL_0) - (\mu_0+\mu^d+\varphi_0)I_0\\
    \dot{J}_0 &= (1-q)(\nu^FF_0 + \nu_0^LL_0) - (\mu_0+\mu^d+\varphi_0)J_0\\
    \dot{S}_1 &= (1-f)\alpha(N_0+N_1) + \sigma_1^FF_1 + \sigma^LL_1 
                +\varphi_1(I_1+J_1) - \lambda_1S_1 - \mu_1S_1\\
    \dot{F}_1 &= gpf\alpha(N_0+N_1) + p\lambda_1S_1 + xp\lambda_1L_1 
                -(\mu_1+\nu^F+\sigma_1^F)F_1\\
    \dot{L}_1 &= (1-gp)f\alpha(N_0+N_1) + (1-p)\lambda_1S_1 - xp\lambda_1L_1 
                -(\mu_1+\nu_1^L+\sigma^L)L_1\\
    \dot{I}_1 &= q(\nu^FF_1 + \nu_1^LL_1) - (\mu_1+\mu^d+\varphi_1)I_1\\
    \dot{J}_1 &= (1-q)(\nu^FF_1 + \nu_1^LL_1) - (\mu_1+\mu^d+\varphi_1)J_1\\
  \end{align*}
  \caption{System of Differential Equations given in the Hill Model.}
  \label{fig:hillEquations}
\end{figure}
\begin{figure}
  \begin{align*}
    \sigma^{L} &= 0.057   &&\text{(Treatment rate for chronic LTBI)}\\
    v^{L}_{0}  &= 0.0014  &&\text{(Progression rate for reactivation in the USB
                                   population)}\\
    v^{L}_{1}  &= 0.0010  &&\text{(Progression rate for reactivation in the FB
                                   population)}\\
    f          &= 0.187   &&\text{(Fraction of FB arrivals with LTBI)}\\
    p          &= 0.103   &&\text{(Fraction of new infections which are acute)}\\
    ARI_{0}    &= 0.00030 &&\text{(2000 Annual Risk of Infection, USB)}\\
    q          &= 0.708   &&\text{(Fraction of infections progressing to infectious
                                   disease)}\\
    g          &= 0.0047  &&\text{(Fraction of FB arrivals with LTBI who are fast
                                   progressors)}\\
    \sigma^{F} &= 0.461   &&\text{(Cumulative fraction of treatment for acute
                                   infection)}\\
    r_{0}      &= 0.667   &&\text{(Fraction of cases due to reactivation in USB
                                   population)}\\
    r_{1}      &= 0.780   &&\text{(Fraction of cases due to reactivation in FB
                                   population)}\\
    \mu^{d}    &= 0.115   &&\text{(Mortality rate due to TB)}\\
    x          &= 0.111   &&\text{(Fraction of re-infected chronic LTBI moving to
                                   acute infection)}\\
    \phi       &= 0.897   &&\text{(Cumulative fraction self-cure and treatment of
                                   active disease)}\\
    e_{0}      &= 0.965   &&\text{(Fraction of preferred contacts within own
                                   population for USB)}\\
    e_{1}      &= 0.985   &&\text{(Fraction of preferred contacts within own
                                  population for FB)}\\
  \end{align*}
  \caption{The relevant constants used in the Basic Hill Model.  }
  \label{fig:basicHillConstants}
\end{figure}

The vector variables $S_0, F_0, L_0, I_0, J_0$ contain the number of US-born
individuals in the S, F, L, I, J compartments respectively, whereas $S_1, F_1,
L_1, I_1, J_1$ contain the number of foreign-born individuals. $N_0$ and $N_1$
are the total populations of US-born and foreign-born individuals. 
\TODO{Is the next bit needed? I think move the stuff below out and the stuff
above back where we introduce the equations.}
The constants $\rho$ and $\alpha$ are birth rates, while $\mu_i$, and $\mu_d$
are death rates. A complete list and descriptions of all constants used in the
model can be found in the Appendix, Section~\ref{subsec:hillConstants}.

\subsection{Additional Tracking Capabilities}
In order to refine the tracking capabilities of the Hill Model, the original
differential equations used to describe TB spread were separated into their
component parts and each part was tracked separately. In effect, each flow rate
between two compartments in the Basic Hill model was made into its own
``virtua'' compartment, thereby enabling computation not only of expected TB
incidence, but also of the source of that incidence. In particular, progressions
into active TB due to activations of LTBI or exogenous infection were tracked,
allowing for sourced incidence data to be generated. These equations allowed the
model to track the sourced number of TB cases, TB deaths, and natural deaths.
Furthermore, the model also tracks the sourced cost on the US HCS. The
immigrating cases of LTBI (acute or chronic) were also tracked. In the case of
intervention testing, the number of cured and untreated cases of entering LTBI
were both tracked. Cured cases of LTBI entering, TB deaths, total TB cases, and
total cost were also tracked discounted at 3\% annually. This discounting was
converted to a continuous differential equation for use within the model. In
general, incidence data is calculated in the same way in the Extended Hill Model
as it was in the Basic Hill Model.

\TODO{Where does this fit in? I'm not seeing why this is here? Maybe in the
results, instead?}
In addition, estimations of the basic reproductive number of FB or USB active,
infectious TB cases were made from a theoretical and an experimental
perspective. Experimental data were calculated by reducing the initial
FB or USB infectious TB populations by one and allowing the
model to progress otherwise as normal. The decreased number of total TB cases
seen details how many infections can be thought to be due to one infected
individual in the given population. In the agent-based model, this \emph{one}
was a true model agent whereas in the compartmental model, this decrease of
\emph{one} was a proportional decrease based on the scaled population level.

From a theoretical perspective, the spread of TB was thought to be a geometric
series. If one infectious individual infects $x$ people annually, over the
course of $N$ years the total number of infections caused by this individual can
be approximated by the geometric series of $N$ terms with rate $x$ and initial
term $1$. In this case, $N = 100$, and the ratios for FB infectious individuals
and USB infectious individuals were obtained from 
\TODO{<+RATIO CITATION COLIN+>.}
\subsection{Economic Modeling}
In order to track the US HCS TB cost, we made the following treatment cost
assumptions. A single case of active TB was assumed to yield a \$14,000 US HCS
cost, charged immediately upon disease contraction. This cost is the weighted
mean of the costs of cases requiring hospitalization (49\%) and cases not
requiring hospitalization (51\%).  These cost estimates were found in 
\TODO{<+DYLAN COST STUDY CITATION+>.} LTBI treatment costs were assumed to be
\$468.00, charged immediately upon successful treatment. These estimates were
calculated based on the cost of a successful treatment, and the typical
adherence and efficacy of said treatment 
\TODO{<+DYLAN COST STUDY CITATION, ADHERENCE CITATION+>.}
It is important to note that the charging assumptions are different for
treatment of active TB as opposed to treatment of latent TB. In particular,
cases of active TB imply a cost immediately upon disease contraction, whereas
latent TB costs are charged only upon successful treatment. The motivation for
this distinction is that active TB treatment is mandated by the US for all known
cases and, further, most treatment costs are incurred at the beginning of the
treatment cycle. As such, charging for every known case, immediately at
contraction is a well motivated assumption. On the other hand, latent TB is only
rarely treated in discovered cases and the costs are more evenly distributed.
Furthermore, the choice to charge only upon successful treatment yields a more
conservative estimate of the success of any intervention analyzed. Finally, it
is worth noting that, though accurate, these estimates of US HCS treatment cost
are not intended to be fixed parameters. The code for this model is freely
available, and as new research amends the necessary costs and new methods of
treatment are developed, these values can be modified and the model re-run to
produce a revised prediction of the US HCS TB cost.

>>>>

% arara: pdflatex
% arara: bibtex
% arara: pdflatex
% arara: pdflatex
\documentclass{amsart}
\usepackage[pdftex]{graphicx}
\usepackage{enumerate,placeins}
\usepackage{standalone}
\usepackage{color}
\usepackage{hyperref}
\usepackage{framed}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows}
\usetikzlibrary{calc}
\graphicspath{{figures/}}
\renewcommand{\(}{\left(}
\renewcommand{\)}{\right)}

\newcommand{\TODO}[1]{\begin{framed}{\huge \color{red} TO DO:}
  #1 \end{framed}}

@@

\TODO{Pick a journal. PLOS?}
\TODO{Format to journal.}
\TODO{Validate against 2000 - 2014, if absolutely necessary. This would require
gathering data, re-running the model, re-performing sensitivity analysis, and
potentially redoing interventions. It may not be necessary as we validated
against the Hill model, but perhaps this should be more stressed.}
\TODO{Change intervention start date to 2014, if absolutely necessary.}
\TODO{Improve the final distributions figure. Is the fact that the deterministic
  hill results are $2+$ standard deviations off the mean acceptable? Where is
  the data for this? Where is the R script to generate these images?}
\TODO{Clean up github.}

\section{Abstract}
An epidemiological and economic model for the spread of tuberculosis in the
United States was developed, extending a model developed in 2012 by the Centers
for Disease Control and Prevention. Various intervention strategies,
particularly those that reduced latent tuberculosis infection (LTBI) among
foreign-born arrivals, were evaluated, and cost per case averted estimates were
found. Curing cases of immigrating LTBI was found to be a highly effective and
cost efficient intervention strategy. Additionally, a population-level,
stochastic, agent-based model was built, which demonstrated the feasibility of
using agent-based modeling in the context of minimal population heterogeneity
and provided statistical characterization of the results of the compartmental
model. The distributions of the final incidence rates projected by the
stochastic model were found to be approximately normal, with means centered at
the results of compartmental model.

\section{Introduction}
As a single infectious agent, tuberculosis (TB) is the second leading cause of
death worldwide, and the World Health Organization (WHO) estimates that roughly
one third of the global population has either active or latent TB
\cite{_who_2013}. This high prevalence does not leave the US untouched; in 2013,
9588 cases of TB were reported in the USA, which is an incidence rate of 30 new
cases per million \cite{miramontes_trends_2013}. In this work, we analyze
various TB intervention strategies aimed at reducing this incidence. In
particular, we evaluate our interventions in the context of a particular goal:
elimination of US TB by 2100. Here, we use elimination to mean that the yearly
incidence of US TB would be lower than 1 new case per million. This goal is
simultaneously ambitious given the prevalence of TB today and also potentially
achievable, as it leaves over 80 years in which to focus on stopping TB.

The majority of TB infections are caused by Mycobacterium tuberculosis, 
an airborne pathogen which primarily affects the lungs and pulmonary system.  
Individuals infected with this bacteria first develop a latent form of TB
referred to as latent tuberculosis infection (LTBI).  These individuals are
asymptomatic and not contagious, and thus, this infection is difficult to detect
and can persist for many years. The latent tuberculosis infection may progress
to active tuberculosis infection (active TB), at which point the patient becomes
symptomatic and contagious. This progression is most likely within the first two
years after contraction, but can happen throughout the rest of the lifespan of
the individual.  Left untreated, individuals with LTBI have a lifelong
probability of approximately 10\% to develop active TB.
\TODO{in total? Or after they've progressed through the first two years? We need
a new source for this, I think. It's not in the logical place of the WHO report.
This also obviously needs to be cited. }
The health risks for individuals with active TB are severe, and without medical
interventions, the 10 year mortality rate for the disease is between 20\% and
70\%, depending on the severity of the case and definition of a positive
diagnosis.
\cite{_who_2013}.

Effective drugs are available to cure almost all cases of LTBI and active TB,
with the exception of drug resistant TB. However, the treatment methods for both
LTBI and active TB are onerous, typically requiring a multi-month drug regiment
of serious antibiotics with potentially harmful side effects, costing an average
of \$14,000 per successful active TB treatment and \$468.00 per successful LTBI
treatment. 
\TODO{The LTBI number here seems off. If I just run
  \href{https://github.com/mmcdermott/disease-modeling/blob/master/in_progress/models/costBenefitAnalysis/hillConstants.R}{\texttt{hillConstants.R}}
  and look at the value of \texttt{Cl}, I get \$700 something. What's going on
  here? Where does the \$468 number come from? What was this supposed to be?}
\TODO{Cite Dylan's paper. Is it the one in zotero?}
Given the high mortality rate for active TB, this treatment is deemed legally
and medically necessary for cases of active TB. However, given the low
progression rate from LTBI to active TB, and the high cost of treatment, this is
often seen as unnecessary for LTBI patients. This becomes especially relevant
when applied to the US immigration program. Before being admitted to the United
States, all potential immigrants are required to take a TB test. If they have
active TB, treatment is required before entry, but if they have LTBI, no
treatment is mandated. As the global incidence for TB is so high, this
contributes a sizable portion of the US LTBI population.

Given that LTBI patients are asymptomatic, many standard TB interventions omit
focusing on LTBI treatment entirely, reasoning that a more aggressive focus on
active TB would serve better in the long run.  However, as LTBI can persist over
a lifetime, such disease control strategies often unintentionally create large
pools of people with LTBI that may or may not eventually develop active TB. This
is especially problematic in the U.S., where our high immigration rate leads to
an even more inflated LTBI population. This leads some to question the
effectiveness of these approaches; in particular, in 2012, Hill, Becerra, and
Castro modeled the spread of TB in the USA and found that controlling the LTBI
population and encouraging treatment of LTBI was essential to TB elimination.
Unfortunately, they also found that reaching elimination was very unlikely, if
not impossible, by 2100.

% I commented out the below because I'm not sure we need it. Thoughts? 
%Disease models like Hill, Becerra, and Castro's provide scientists and
%policy makers some of their only tools when predicting intervention
%effectiveness and cost, and are thus of prime importance when analyzing disease
%spread and planning intervention strategies. However, forming such models and
%extracting meaningful information about the underlying disease spread from their
%results can often be very difficult. The choice of appropriate modeling
%techniques is therefore essential in providing the most information possible
%about a disease structure.

Hill, Becerra, and Castro's model (henceforth called the Hill model) is a
compartmental model, meaning the model splits the population apart into disjoint
``compartments,'' each of which represents a particular health state. The Hill
model utilized a system of 10 compartments split across two sub-populations:
US-born individuals (USB) and foreign born individuals (FB).  Each
sub-population was split among five health states representing various subtypes
of active or latent TB. The flow rates between these compartments were given by
deterministic differential equations acting on a continuous population.
Though this model is extremely effective at modeling TB behavior that agrees on
average with real population data, it lacks tracking capabilities that
would shed more light on the underlying cause of US TB incidence.
Furthermore, it is also limited as it fails to provide any economic data
regarding the cost to the US health care system due to TB (HCS cost) with our
without various interventions. To address these shortcomings, we produced an
extended version of the Hill model, which we refer to as the extended Hill
model, which analyzes the same compartments as the Hill model, but also tracks
the economic impact of TB and tracks the source of all infections and costs
incurred, thus enabling much more detailed qualitative analysis of US TB
dynamics.

Though deterministic compartmental models like the Hill model are well
established in epidemiology, they also fail to capture the natural stochasticity
of real world disease spread, and thus cannot provide any statistical
characterization of their predicted outcomes. To understand these statistical
factors for the extended Hill model, we also produced a stochastic, agent based,
population level model of US TB dynamics that agreed with the Hill model on
average, but provided meaningful statistical parameters regarding the final
distributions of each of the various health states. In addition to providing
statistical verification of the extended Hill results, this model also
demonstrates that agent based modeling can be an effective strategy even in
cases where the population size is very large, as the model ran quickly
even on standard hardware.  This is especially promising as agent based models
are much better equipped to model disease spread in highly heterogeneous
populations, such as in cases of drug resistance or comorbidity, both of which
are very relevant with TB and present excellent opportunities for model
extension.

Using both the extended Hill and the agent based model, we have demonstrated
that LTBI treatment is essential in controlling US TB dynamics and reaching the
2100 elimination goal, if that is possible at all. Furthermore, we have found
that treating cases of immigrating LTBI in the US is critical to our overall TB
control strategy and is cost effective compared to other interventions. In this
paper, we first document the theoretical and implementation details of both the
basic and extended Hill models, and then we provide significant results from
these models. This work has practical implications for TB public health policy
in the United States, and it also demonstrates the feasibility of using agent
based models for large-scale applications in epidemiology.

\section{The Hill Model}
\label{sec:hillModel}
A flowchart representation of the Hill model is shown in
Figure~\ref{fig:hillModelSchematic}.

\begin{figure}[h]
  \begin{center}
    \includestandalone[width=\textwidth]{hillFlowChart}
    %\includegraphics[scale=0.25]{figures/HillModelFlowChart.png}
  \end{center}
  \caption{Schematic of the Hill model. Each compartment represents a different
    possible TB health state for every US-born or FB individual, and arrows
    between different compartments represent possible transitions between
    states. Individuals also leave the model from all compartments due to
    natural death, which is left out of the figure for clarity. Figure adapted
    from \cite{hill_modelling_2012}}
  \label{fig:hillModelSchematic}
\end{figure}

The majority of USB and FB individuals fall into the susceptible ($S$) category,
which includes everyone who is uninfected and has not been exposed to TB, or
those who have been successfully cured of the disease. Individuals in the $S$
category flow with some rate into an LTBI category, simulating exposure to TB.
Because real LTBI patients have a higher risk of developing active TB within two
years of exposure, the Hill model splits the LTBI population among two LTBI
compartments: compartment $F$, or acute LTBI, representing those who progress
within the first two years of catching the disease, and compartment $L$, or
chronic LTBI, representing those who do not progress within the first two years.
Both LTBI compartments feed into two active TB compartments, $I$ and $J$,
representing infectious and non-infectious active TB, respectively. To respect
the varying rates of LTBI activation, individuals in the Acute LTBI compartment
have a much higher risk of developing active TB than those in the Chronic LTBI
compartment. Individuals in the Chronic LTBI compartment may also be exogenously
re-infected and transition to the Acute LTBI compartment
\cite{hill_modelling_2012}.

Individuals in both active TB compartments have an increased risk of death from
TB infection, but only individuals in the infectious TB compartment affect the
infection flow rates. In addition, individuals in all of the infected
compartments ($F$, $L$, $I$, $J$) may be treated or self-cure themselves of
their respective TB health condition. However treatment or self-cure from TB
does not grant immunity, and all healthy individuals are grouped in the
susceptible compartment and may be re-infected at a later time
\cite{hill_modelling_2012}.

The rates that individuals move into and between compartments in the Hill model
are determined by a system of deterministic differential equations. This system
of equations, and the definitions of the constants used in the Hill model, are listed in
Appendix~\ref{appendix:hillEqs} \cite{hill_modelling_2012}. These equations
provide accurate estimations of the sizes of each compartment at any point in
time, but can provide little insight into TB dynamics beyond that, as they offer
no infection source information. Furthermore, it is important to note once again
that though real-world disease dynamics are stochastic, the Hill model is a
completely deterministic approximation and does not provide estimates for the
distributions or standard errors of predicted disease rates. We addressed each
of these issues with the extended Hill model and the agent based Hill model.

\section{The Extended Hill Model}
\label{sec:extendHill}
In order to refine the tracking capabilities of the Hill model, the original
differential equations used to describe TB spread were separated into their
component parts and each part was tracked separately. In effect, each flow rate
between two compartments in the basic Hill model was made into its own
``virtual'' compartment, which is tracked the same as a real compartment by the
model but does not represent a particular health state. This enabled computation
not only of expected TB cases and deaths, but also the source of those cases and
deaths. A schematic for this model is shown in Figure~\ref{fig:extendedHill}.
Immigrating cases of LTBI (acute or chronic) were also tracked. In the case of
intervention testing, the number of cured and untreated cases of entering LTBI
were both tracked.

\begin{figure}[h]
  \centering
  \includestandalone[width=\textwidth]{extendedHillFlowChart}
  \caption{A schematic of the extended Hill model. True compartments are
    illustrated in red, with flow rates illustrated with black arrows and flow
    tracking compartments illustrated in green with dashed edges.
  }
  \label{fig:extendedHill}
\end{figure}

The model was also extended to provide similarly sourced US health care system
(HCS) TB costs. To do so, we made the following treatment cost assumptions. A
single case of active TB was assumed to yield a \$14,000 US HCS cost, charged
immediately upon disease contraction. This cost is the weighted mean of the
costs of cases requiring hospitalization (49\%) and cases not requiring
hospitalization (51\%).  These cost estimates were found in \TODO{<+DYLAN COST
STUDY CITATION+>.} LTBI treatment costs were assumed to be \$468.00, charged
immediately upon successful treatment. These estimates were calculated based on
the cost of a successful treatment, and the typical adherence and efficacy of
said treatment \TODO{<+DYLAN COST STUDY CITATION, ADHERENCE CITATION+>.} Note
that active TB treatment costs are charged immediately upon disease contraction,
whereas LTBI treatment costs are charged only upon successful treatment. The
motivation for this distinction is that active TB treatment is mandated by the
US for all known cases and, further, most treatment costs are incurred at the
beginning of the treatment cycle. As such, charging for every known case,
immediately at contraction is a well motivated simplification. On the other
hand, latent TB is only rarely treated in discovered cases and the costs are
more evenly distributed. Furthermore, the choice to charge only upon successful
treatment yields a more conservative estimate of the success of any intervention
analyzed, as it will always underestimate the total US HCS TB cost burden.
Finally, it is worth noting that these estimates of US HCS treatment cost are
not intended to be fixed parameters. The code for this model is freely
available, and as new research amends the necessary costs and new methods of
treatment are developed, these values can be modified and the model re-run to
produce a revised prediction of the US HCS TB cost. In addition to US HCS cost,
the system was implemented so as to also track projected intervention
implementation cost, given user-inputted parameters relating to various possible
intervention cost strategies.  Cured cases of LTBI entering, TB deaths, total TB
cases, and total US HCS cost were also tracked discounted at 3\% annually for
cost benefit analysis. This discounting was converted to a continuous
differential equation for use within the model.

The extended Hill model was implemented as a series of differential equations in
\texttt{R}, and solved using the \texttt{lsoda} system with a timestep of $0.8$.
Extensive sensitivity analysis was performed on this parameter and it was found
that decreasing the timestep did not seriously affect the model outcomes.  A
relevant subset of the extended Hill equations and a link to the full model code
and documentation can be found in Appendix~\ref{appendix:hillEqs}. Historical
data from 2000-2008 was used to initialize the model, which was then run up to
the year 2100.

In order to understand these results statistically, a stochastic, agent based
model of TB spread in the US was implemented. This model was programmed in
\texttt{C++}, and probabilities of agent transition between various health
states were computed given rates in the hill model and a variety of integration
approximations. The model maintained individual records for every individual
with LTBI or active TB. Susceptible individuals were not modeled as agents;
instead, a binomial distribution was used to probabilistically determine the
number of new infections each time step. We ran the model with default parameter
values for 2160 runs, with time step of $0.01$.  Final population health
statistics were recorded for each run, and these data were analyzed in $R$. The
agent based model also tracks deaths, total TB cases, and infection sources.
These data were consistently normal and matched the deterministic model closely
in mean, providing statistical validation of using the deterministic model
results in intervention analysis and policy planning without more advanced
statistical analysis.

Further implementations of the agent based model were made that deviated from
the basic Hill model. Specifically, the acute latent compartment in the basic
Hill model is a vestigial necessity of compartmental modeling and is not
reflective of the biology of tuberculosis. As such, an agent based model was
implemented with one LTBI health state that relied on time-dependent
probabilities to ensure that individuals transferred within the first two years
with higher probability.  Results from this model were also normal and did not
differ significantly from the strict Hill model. This also demonstrates the
effectiveness of using agent based modeling at population level scale for
general epidemiological studies. Agent based models are much better equipped to
model population heterogeneity, drug resistance, or comorbidity, but are
typically thought to be too computationally intensive to perform on large
populations. These results show that with modern hardware, agent based modeling
can be used even with a population of millions of agents.

\section{Results}

\subsection{Basic Population Breakdown}
The additional tracking capabilities of the extended Hill model offered several
key insights into US TB dynamics. Figure~\ref{fig:incPlotSourced} shows the
yearly incidence of US TB, broken down by infection source. Recall that
infections due to acute LTBI, in the Hill model, represent cases of a novel
infection, whereas infections due to chronic LTBI represent the activation of a
long standing latent infection.  Figure~\ref{fig:incPlotSourced} demonstrates
that the majority of the US TB load is driven by activations of FB LTBI,
followed by USB LTBI activations. Though contact investigation and aggressive TB
treatment programs help prevent increasing the LTBI population, as active TB
cases can also lead to long lasting LTBI cases, this suggests that a greater
focus on LTBI treatment would potentially have a very large impact on US TB
rates. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill model.}
  \label{fig:incPlotSourced}
\end{figure}

Figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to active TB. One can see that in this plot, roughly half
of the US TB HCS costs are due to activations of LTBI. Note that, again, in this
figure it is appropriate to think of the costs due to acute LTBI as costs due to
exogenous re-inection, whereas costs due to chronic LTBI are activations of long
standing latent infections. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
